---
title: "Arms trade: US spending"
subtitle: "Mentions of top 100 arms companies in US defence contracts"
author: "James Goldie, 360info"
date: "2022-07-15"
code-fold: true
theme: style/article.scss
---

```{r}
library(tidyverse)
library(arrow)
```

Information on US contracts is published at [usaspending.gov](https://usaspending.gov). We'd like to see if we can identify contracts with the SIPRI top 100 arms companies.

The full database is a little heavy for our purposes, so we're going to download [custom account data](https://www.usaspending.gov/download_center/custom_account_data) for the Department of Defence each year in turn.

::: {.column-margin}
These are current as of July 18, 2022, but you can [download newer versions of the custom account data](https://www.usaspending.gov/download_center/custom_account_data) using the following options:

**Budget Function**

National Defense

**Budget Sub-Function**

All

**Agency**

All

**Federal Account**

Federal account not selected

**Account Level**

Treasury Account

**Disaster Emergency Fund Codes**

—

**File Submission Type**

Account Breakdown by Program Activity & Object Class (File B), Account Breakdown by Award* (File C)

**Fiscal Year**

—
:::

This data has been backed up to the 360 Shared Drive in Parquet format, so we'll use the [`{arrow}` package](https://arrow.apache.org/docs/r/reference/write_parquet.html) to retrieve it.




```{r}
acc %>%
  filter(str_detect(recipient_name, coll("raytheon", ignore_case = TRUE))) ->
acc_raytheon

acc %>%
  write_parquet(
    "accounts-FY2022.gz.parquet",
    compression = "gzip",
    compression_level = 5)
```